A High-Order Discontinuous Galerkin Solver with Dynamic Adaptive Mesh Refinement to Simulate Cloud Formation Processes

### 2019-05-04 writing/paper/paper.tex
Add version to be submitted to PPAM, final (?).

\begin{document}
\maketitle 
\begin{abstract}
  We present a high-order discontinuous Galerkin (\dg{}) solver of the compressible Navier Stokes equations for cloud formation processes.
  The scheme exploits an underlying parallelized implementation of the \aderdg{} method with dynamic adaptive mesh refinement. 
  We improve our method by a \pde{}-independent general refinement criterion, based on the local total variation of the numerical solution.
  Our generic scheme shows competitive results for both classical \textsc{cfd} and stratified scenarios.
  While established methods use tailored numerics towards the specific simulation, our scheme works scenario independent.
  All together our method can be seen as a perfect candidate for large scale cloud simulation runs on current and future super computers.

@@

As a final example of standard flow scenarios, we consider the lid-driven cavity flow where the fluid is initially at rest, with $\Qrho = 1$ and $ \pressure(\bm{x}) = \nicefrac{100}{\gamma}$.
We consider a domain of size $\SI{1}{m} \times \SI{1}{\m}$ which is surrounded by no-slip walls.
The flow is driven entirely by the upper wall which has a velocity of $v_x = \SI{1}{\m/\s}$.
The simulation runs for $\SI{10}{\s}$
Again, our results (\cref{fig:cavity-flow}) have an excellent agreement with the reference solution of~\cite{ghia1982high}.
We used an \aderdg{}-method of order $3$ with a mesh of size $27^2$.

With  the constants
\begin{equation}\label{eq:atmosphere-constants}
    \gamma = 1.4 ,\quad \Pr =  0.71 ,\quad R = 287.058 ,\quad p_0 = \SI{10e5}{\Pa}, \quad g = \SI{9.8}{m/s^2},
\end{equation}
all following stratified flow scenarios are described in terms of the potential temperature
\begin{equation}
  \potT = T \left( \frac{p_o}{p} \right)^{R/c_p},
\end{equation}
with reference pressure $p_o$~\cite{muller2010adaptive,giraldo2008study}.
%
We compute the initial background density and pressure by inserting the assumption of a constant background energy in \cref{eq:hydrostatic-balance}.
The background atmosphere is then perturbed.
We set the density and energy at the boundary such that it corresponds to the background atmosphere.
Furthermore, to ensure that the atmosphere stays in hydrostatic balance, we need to impose the viscous heat flux
\begin{equation}
  \label{eq:atmosphere-bc}
  \viscFlux_{\QE} = \kappa \pdv{\overline{T}}{z}.% =
\end{equation}
at the boundary.
In this equation, $\overline{T}(z)$ is the background temperature at position $z$, which can be computed from \cref{eq:hydrostatic-balance,eq:eos}~\cite{giraldo2008study}.

Our first scenario is the colliding bubbles scenario~\cite{muller2010adaptive}.
We use perturbations of the form
\begin{equation}
  \label{eq:bubbles-pertubation}
  \pertubationPotT =
  \begin{cases}
    A & r \leq a, \\
    A \exp \left( - \frac{(r-a)^2}{s^2} \right) & r > a,
    \end{cases}
\end{equation}
where $s$ is the decay rate and $r$  is the radius to the center~\pcref{eq:radius}:
\begin{equation}
  \label{eq:radius}
  r^2 = \Vert \bm{x} - \bm{x_c} \Vert_2
\end{equation}
i.e., $r$ denotes the Euclidean distance between the spatial positions $\bm{x} = (x, z)$ and the center of a bubble $\bm{x_c} = (x_c, z_c)$ -- for three-dimensional scenarios $\bm{x}$ and $\bm{x_c}$ also contain a $y$ coordinate.

We have two bubbles, with constants
\begin{equation}
  \label{eq:bubbles-values}
\begin{alignedat}{6}
  & \text{warm:} \qquad && A = \SI{0.5}{\K}, \quad&& a = \SI{150}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{500}{\m,} \quad&& z_c = \SI{300}{\m},\\
  & \text{cold:} \qquad && A = \SI{-0.15}{\K}, \quad&& a = \SI{0}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{560}{\m}, \quad&& z_c = \SI{640}{\m}.
  \end{alignedat}
\end{equation}
Similar to~\cite{muller2010adaptive}, we use a constant viscosity of $\mu = 0.01$ to regularize the solution.
Note that we use a different implementation of viscosity than~\cite{muller2010adaptive}.
Hence, it is difficult to compare the parametrization directly.
We ran this scenario twice: once without \amr{} and a mesh of size $\SI{1000/81}{\m} = \SI{12.35}{\m}$ and 
once with \amr{} with two adaptive refinement levels and parameters $\Trefine = 2.5$ and $\Tdelete = -0.5$.
We specialize the \amr{}-criterion~\pcref{eq:refinement-criterion} to our stratified flows by using the potential temperature.
This resulted in a mesh with cell-size lengths of approx.\ \SI{111.1}{\m}, \SI{37.04}{\m}, and \SI{12.34}{\m}.
The resulting mesh can be observed in \cref{fig:two-bubbles-ader}.
We observe that the $L_2$ difference between the potential temperature of the \amr{} run, which uses 1953 cells, and the one of the fully refined run with 6561 cells, is only $1.87$.
The relative error is \num{6.69e-6}.
We further emphasize that our \amr{}-criterion accurately tracks the position of the edges of cloud instead of only its position.
This is the main advantage of our gradient-based method in contrast to methods working directly with the value of the solution, as for example~\cite{muller2010adaptive}.
Overall, our result for this benchmark shows an excellent agreement to the previous solutions of~\cite{muller2010adaptive}.

@@

For our second scenario, the cosine bubble, we use a perturbation of the form
\begin{align}
  \label{eq:cos-pertubation}
  \pertubationPotT &= \begin{cases}
    \nicefrac{A}{2} \left[ 1 + \cos(\pi r) \right] & r \leq a, \\
    0 & r > a,
    \end{cases}
\end{align}
where $A$ denotes the maximal perturbation and $a$ is the size of the bubble.
We use the constants
\begin{equation}\label{eq:cosine-bubble}
  A = \SI{0.5}{\K} \quad a = \SI{250}{\m} \quad x_c = \SI{500}{\m} \quad z_c = \SI{350}{\m}.
\end{equation}
For the three-dimensional bubble, we set $y_c = x_c = \SI{500}{\m}$ and lower the bubble to $z_c = \SI{260}{\m}$.
This corresponds to the parameters used in~\cite{kelly2012continuous}\footnote{%
We found that the parameters presented in the manuscript of~\cite{kelly2012continuous} only agree with the results, if we use the same parameters as for 2D simulations.}.
For the 2D case, we use a constant viscosity of $\mu = 0.01$ and an \aderdg{}-method of order 5 with two levels of dynamic \amr{}, resulting again in cell sizes of roughly $\SI{111.1}{\m}, \SI{37.04}{\m}, \SI{12.34}{\m}$.
We use slightly different \amr{} parameters of $\Trefine = 1.5$ and $\Tdelete = -0.5$ and let the simulation run for \SI{600}{\s}.
Note that, as seen in \cref{fig:cosine-2d}, our \amr{}-criterion tracks the wavefront of the cloud accurately.
This result shows an excellent agreement to the ones achieved in~\cite{giraldo2008study,muller2010adaptive}.

For the 3D case, we use an \aderdg{}-scheme of order 3 with a static mesh with cell sizes of \SI{40}{\m}. and a shorter simulation duration of \SI{400}{\s}.
Due to the relatively coarse resolution and the hence increased aliasing errors, we need to increase the viscosity to $\mu = 0.02$.
This corresponds to a larger amount of smoothing.
Our results (\cref{fig:cosine-3d}) capture the overall dynamics of the scenario well, similar to the reference solution of~\cite{kelly2012continuous}.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.90\textwidth]{paper_two_bubbles}
  \caption{\label{fig:two-bubbles-ader}Colliding Bubbles \aderdg{}. Contour values for potential temperature perturbation are $-0.05, 0.05, 0.1, \ldots 0.45$.}
\end{figure}

\begin{figure}[tb]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{paper_cosine_bubble}
    \caption{\label{fig:cosine-2d}2D Cosine Bubble.}%todo: Contour lines?
  \end{subfigure}~%
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{paper_cosine_bubble_3d}
    \caption{3D Cosine Bubble}
  \end{subfigure}
  \caption{\label{fig:cosine-3d}Cosine Bubble Scenarios}
  \label{fig:cosine-bubbles-results}
\end{figure}

>>>>

\begin{document}
\maketitle 
\begin{abstract}
  We present a high-order discontinuous Galerkin (\dg{}) solver of the compressible Navier Stokes equations for cloud formation processes.
  The scheme exploits an underlying parallelized implementation of the \aderdg{} method with dynamic adaptive mesh refinement. 
  We improve our method by a \pde{}-independent general refinement criterion, based on the local total variation of the numerical solution.
  Our generic scheme shows competitive results for both classical \textsc{cfd} and stratified scenarios.
  While established methods use numerics tailored towards the specific simulation, our scheme works scenario independent.
  All together, our method can be seen as a perfect candidate for large scale cloud simulation runs on current and future super computers.

@@

As a final example of standard flow scenarios, we consider the lid-driven cavity flow where the fluid is initially at rest, with $\Qrho = 1$ and $ \pressure(\bm{x}) = \nicefrac{100}{\gamma}$.
We consider a domain of size $\SI{1}{m} \times \SI{1}{\m}$ which is surrounded by no-slip walls.
The flow is driven entirely by the upper wall which has a velocity of $v_x = \SI{1}{\m/\s}$.
The simulation runs for $\SI{10}{\s}$.
Again, our results (\cref{fig:cavity-flow}) have an excellent agreement with the reference solution of~\cite{ghia1982high}.
We used an \aderdg{}-method of order $3$ with a mesh of size $27^2$.

With  the constants
\begin{equation}\label{eq:atmosphere-constants}
    \gamma = 1.4 ,\quad \Pr =  0.71 ,\quad R = 287.058 ,\quad p_0 = \SI{10e5}{\Pa}, \quad g = \SI{9.8}{m/s^2},
\end{equation}
all following stratified flow scenarios are described in terms of the potential temperature
\begin{equation}
  \potT = T \left( \frac{p_0}{p} \right)^{R/c_p},
\end{equation}
with reference pressure $p_0$~\cite{muller2010adaptive,giraldo2008study}.
%
We compute the initial background density and pressure by inserting the assumption of a constant background energy in \cref{eq:hydrostatic-balance}.
The background atmosphere is then perturbed.
We set the density and energy at the boundary such that it corresponds to the background atmosphere.
Furthermore, to ensure that the atmosphere stays in hydrostatic balance, we need to impose the viscous heat flux
\begin{equation}
  \label{eq:atmosphere-bc}
  \viscFlux_{\QE} = \kappa \pdv{\overline{T}}{z}.% =
\end{equation}
at the boundary~\cite{giraldo2008study}.
In this equation, $\overline{T}(z)$ is the background temperature at position $z$, which can be computed from \cref{eq:hydrostatic-balance,eq:eos}.

Our first scenario is the colliding bubbles scenario~\cite{muller2010adaptive}.
We use perturbations of the form
\begin{equation}
  \label{eq:bubbles-pertubation}
  \pertubationPotT =
  \begin{cases}
    A & r \leq a, \\
    A \exp \left( - \frac{(r-a)^2}{s^2} \right) & r > a,
    \end{cases}
\end{equation}
where $s$ is the decay rate and $r$  is the radius to the center
\begin{equation}
  \label{eq:radius}
  r^2 = \Vert \bm{x} - \bm{x_c} \Vert_2,
\end{equation}
i.e., $r$ denotes the Euclidean distance between the spatial positions $\bm{x} = (x, z)$ and the center of a bubble $\bm{x_c} = (x_c, z_c)$ -- for three-dimensional scenarios $\bm{x}$ and $\bm{x_c}$ also contain a $y$ coordinate.

We have two bubbles, with constants
\begin{equation}
  \label{eq:bubbles-values}
\begin{alignedat}{6}
  & \text{warm:} \qquad && A = \SI{0.5}{\K}, \quad&& a = \SI{150}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{500}{\m,} \quad&& z_c = \SI{300}{\m},\\
  & \text{cold:} \qquad && A = \SI{-0.15}{\K}, \quad&& a = \SI{0}{\m}, \quad&& s = \SI{50}{\m}, \quad&& x_c = \SI{560}{\m}, \quad&& z_c = \SI{640}{\m}.
  \end{alignedat}
\end{equation}
Similar to~\cite{muller2010adaptive}, we use a constant viscosity of $\mu = 0.01$ to regularize the solution.
Note that we use a different implementation of viscosity than~\cite{muller2010adaptive}.
Hence, it is difficult to compare the parametrization directly.
We ran this scenario twice: once without \amr{} and a mesh of size $\SI{1000/81}{\m} = \SI{12.35}{\m}$ and 
once with \amr{} with two adaptive refinement levels and parameters $\Trefine = 2.5$ and $\Tdelete = -0.5$.
We specialize the \amr{}-criterion~\pcref{eq:refinement-criterion} to our stratified flows by using the potential temperature.
This resulted in a mesh with cell-size lengths of approx.\ \SI{111.1}{\m}, \SI{37.04}{\m}, and \SI{12.34}{\m}.
The resulting mesh can be seen in \cref{fig:two-bubbles-ader}.
We observe that the $L_2$ difference between the potential temperature of the \amr{} run, which uses 1953 cells, and the one of the fully refined run with 6561 cells, is only $1.87$.
The relative error is \num{6.69e-6}.
We further emphasize that our \amr{}-criterion accurately tracks the position of the edges of the cloud instead of only its position.
This is the main advantage of our gradient-based method in contrast to methods working directly with the value of the solution, as for example~\cite{muller2010adaptive}.
Overall, our result for this benchmark shows an excellent agreement to the previous solutions of~\cite{muller2010adaptive}.

@@

For our second scenario, the cosine bubble, we use a perturbation of the form
\begin{align}
  \label{eq:cos-pertubation}
  \pertubationPotT &= \begin{cases}
    \nicefrac{A}{2} \left[ 1 + \cos(\pi r) \right] & r \leq a, \\
    0 & r > a,
    \end{cases}
\end{align}
where $A$ denotes the maximal perturbation and $a$ is the size of the bubble.
We use the constants
\begin{equation}\label{eq:cosine-bubble}
  A = \SI{0.5}{\K} \quad a = \SI{250}{\m} \quad x_c = \SI{500}{\m} \quad z_c = \SI{350}{\m}.
\end{equation}
For the three-dimensional bubble, we set $y_c = x_c = \SI{500}{\m}$.
This corresponds to the parameters used in~\cite{kelly2012continuous}\footnote{%
We found that the parameters presented in the manuscript of~\cite{kelly2012continuous} only agree with the results, if we use the same parameters as for 2D simulations.}.
For the 2D case, we use a constant viscosity of $\mu = 0.01$ and an \aderdg{}-method of order 5 with two levels of dynamic \amr{}, resulting again in cell sizes of roughly $\SI{111.1}{\m}, \SI{37.04}{\m}, \SI{12.34}{\m}$.
We use slightly different \amr{} parameters of $\Trefine = 1.5$ and $\Tdelete = -0.5$ and let the simulation run for \SI{600}{\s}.
Note that, as seen in \cref{fig:cosine-2d}, our \amr{}-criterion tracks the wavefront of the cloud accurately.
This result shows an excellent agreement to the ones achieved in~\cite{giraldo2008study,muller2010adaptive}.

For the 3D case, we use an \aderdg{}-scheme of order 3 with a static mesh with cell sizes of \SI{40}{\m} and a shorter simulation duration of \SI{400}{\s}.
Due to the relatively coarse resolution and the hence increased aliasing errors, we need to increase the viscosity to $\mu = 0.05$.
This corresponds to a larger amount of smoothing.
Our results (\cref{fig:cosine-3d}) capture the dynamics of the scenario well and agree with the reference solution of~\cite{kelly2012continuous}.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.90\textwidth]{paper_two_bubbles}
  \caption{\label{fig:two-bubbles-ader}Colliding Bubbles with \aderdg{}. Contour values for potential temperature perturbation are $-0.05, 0.05, 0.1, \ldots 0.45$.}
\end{figure}

\begin{figure}[tb]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0 23cm 0},clip]{paper_cosine_bubble_collage}
    %\includegraphics[width=\textwidth]{paper_cosine_bubble}
    \caption{\label{fig:cosine-2d}2D Cosine Bubble.}%todo: Contour lines?
  \end{subfigure}~%
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, trim={23cm 0 0cm 0},clip]{paper_cosine_bubble_collage}
    %\includegraphics[width=\textwidth]{paper_cosine_bubble_3d}
    \caption{\label{fig:cosine-3d}3D Cosine Bubble}
  \end{subfigure}
  \caption{\label{fig:cosine-bubbles-results}Cosine Bubble Scenario}
\end{figure}
